---
title: "Public Transportation Acces and Reach - Analysis"
output: html_document
date: "2024-06-09"
---

The analysis of the differences in public transportation access and reach is conducted in this markdown. For more than the comments given, a complete walk through and decsription can be found in the README.md file in the repository. 

# Set up: Package Installation and Data Load 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load("tidyverse", "sf", "mapview", "ggplot2", "geodata", "emmeans", "webshot2")
```

```{r}
#the sf object with distances travelled within each time-limit
dist_sf <- st_read("../in/preprocessed/GTFS_prep/distancesALL.shp")
#the stops_sf object 
stops_sf <- st_read("../in/preprocessed/GTFS_prep/stops.shp")
EDU <- st_read("../in/preprocessed/education/EDU.shp") #load in EDU data
```

Polygons of all municipalities are loaded in from GADM. It is converted to an sf object and transformed to crs = 25832. Only the region name (REG_NAME) and municipality name (MUN_NAME) (and the geometry column) are kept in the sf object
```{r}
##retreieve municipality boundaries 
mun <- gadm(country = "DNK", level = 2, path = "../in/preprocessed/")
mun_sf <- mun %>% 
  st_as_sf() %>% 
  st_transform(crs= 25832) %>% 
  dplyr::select("NAME_1", "NAME_2") 
```

# Data Inspection and Summary statiscs 
Add the status column to dist_sf and make predictor variables ordered factors
```{r}
dist_sf$STATUS <- EDU$STATUS[dist_sf$EDU_i] #add status col to sist 
dist_sf <- dist_sf[c(2,12,10,3,6,4,5,1,7:9)]

dist_sf$STATUS <- ordered(dist_sf$STATUS)
dist_sf$limit <- ordered(dist_sf$limit , 
                         levels = c("5 minute", "20 minute", "45 minute"))
```


## Visual Inspection and data Transformation
The visual inspection reveals a skew in the distance data, driven by the many routes not reaching a stop within the maximum 45 minutes. To be able to log transform the data, these rows with dist = 0 are subsetted. 
```{r}
##log transforming the distance data as it is very skewed
png("../out/total_distances.png")
hist(dist_sf$distanc,
     main = "Distances Travelled Total", 
     xlab = "Distance (meters)")
dev.off()

dist_sf0 <- dist_sf[dist_sf$distanc ==0,] #subset dist = 0 values 
dist_sf <- dist_sf[!dist_sf$distanc ==0,] #subset dist = 0 values 

dist_sf$distanc_log <- log(dist_sf$distanc) #log transdform 

png("../out/log_distances.png")
hist(dist_sf$distanc_log, 
     main = "Log Transformed distances Travelled", 
     xlab = "Log-Transformed Distance")
dev.off()
```

## Summary Statistics for Distances > 0 meters 
The summary statiscs are calculated based on the distance in meters, and saved to the out folder 
```{r}
columns <- c("status", "limit", "mean", "sd", "n")
dist_summary = data.frame(matrix(nrow = 8, ncol = length(columns))) 
colnames(dist_summary) <- columns
dist_summary$status <- c("Old", "Old", "Old", "Old", "New", "New", "New", "New")
dist_summary$limit <- c("total", "5 minute", "20 minute", "45 minute", 
                        "total", "5 minute", "20 minute", "45 minute")
dist_summary$mean <- c(mean(dist_sf$distanc[dist_sf$STATUS == "Old"]), 
                       mean(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "5 minute"]), 
                       mean(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "20 minute"]), 
                       mean(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "45 minute"]), 
                       mean(dist_sf$distanc[dist_sf$STATUS == "New"]), 
                       mean(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "5 minute"]), 
                       mean(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "20 minute"]), 
                       mean(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "45 minute"]))
dist_summary$sd <- c(sd(dist_sf$distanc[dist_sf$STATUS == "Old"]), 
                     sd(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "5 minute"]), 
                     sd(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "20 minute"]),
                     sd(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "45 minute"]),
                     sd(dist_sf$distanc[dist_sf$STATUS == "New"]), 
                     sd(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "5 minute"]), 
                     sd(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "20 minute"]),
                     sd(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "45 minute"]))
dist_summary$n <- c(length(dist_sf$distanc[dist_sf$STATUS == "Old"]), 
                     length(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "5 minute"]), 
                     length(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "20 minute"]),
                     length(dist_sf$distanc[dist_sf$STATUS == "Old" & dist_sf$limit == "45 minute"]),
                     length(dist_sf$distanc[dist_sf$STATUS == "New"]), 
                     length(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "5 minute"]), 
                     length(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "20 minute"]),
                     length(dist_sf$distanc[dist_sf$STATUS == "New" & dist_sf$limit == "45 minute"]))

dist_summary[nrow(dist_summary)+1,] <- c("total", "total", 
                                         mean(dist_sf$distanc),
                                         sd(dist_sf$distanc), 
                                         length(dist_sf$distanc))
##mean, sd and n numeric
dist_summary$mean <- as.numeric(dist_summary$mean)
dist_summary$sd <- as.numeric(dist_summary$sd)
dist_summary$n <- as.numeric(dist_summary$n)

##save to csv. 
write.csv(dist_summary, file = "../out/summary_stats.csv")
```

## Summarise Distances < 0 meters.
```{r}
#short summary of distance = 0

dist0_summary <- dist_sf0 %>% 
  as.data.frame() %>% #to be able to view summary stats wihtout the geometry column
  select(limit, STATUS) %>% 
  summarise(n = n(), 
            .by = c(STATUS, limit)) 

#extract percentage of observations of the entire set that has dist = 0
n0 <- dist0_summary$n
totaln <-n0 + as.numeric(c(
  dist_summary$n[dist_summary$status == "Old" & dist_summary$limit =="45 minute"],
  dist_summary$n[dist_summary$status == "Old" & dist_summary$limit =="5 minute"], 
  dist_summary$n[dist_summary$status == "Old" & dist_summary$limit =="20 minute"], 
  dist_summary$n[dist_summary$status == "New" & dist_summary$limit =="45 minute"], 
  dist_summary$n[dist_summary$status == "New" & dist_summary$limit =="20 minute"], 
  dist_summary$n[dist_summary$status == "New" & dist_summary$limit =="5 minute"]))

dist0_summary$perc <-  n0/totaln*100
write.csv(dist0_summary, file = "../out/summary_dist0.csv")

png("../out/map_dist0.png")
plot(st_geometry(mun_sf), main = "Programmes and Stops, where distance = 0"); plot(stops_sf$geometry[stops_sf$stop_id %in% unique(dist_sf0$stop_id)], col = "red", pch = 19, add = T);plot(EDU$geometry[unique(dist_sf0$EDU_i)], col = sf.colors(2), pch = 4, add= T)
dev.off()

```
## Box-and density plots
```{r}
ggplot(dist_sf, aes(x=distanc_log, group = STATUS, fill = STATUS))+
  geom_boxplot(alpha = 0.5)+
  facet_wrap(~limit, nrow = 3)

ggplot(dist_sf, 
       aes(x=distanc_log, group = STATUS, col = STATUS, fill = STATUS))+
  geom_density(alpha = 0.9)+
  facet_wrap(~limit, nrow = 3)
```

# Linear Regression on Distance
```{r}
png("../out/perc_dist0.png")
ggplot(dist0_summary)+
  geom_col(aes(x = limit, y = perc, color = STATUS, fill = STATUS), 
           position = position_dodge(0.8), width = 0.7, alpha = 0.7) 
dev.off()

m1 <- (lm(distanc_log ~ STATUS + limit, data = dist_sf))
summary(m1)
png("../out/logdist_programmes.png")
emmip(m1, STATUS ~ limit, CIs = TRUE, ylab= "log distance", xlab = "") 
dev.off()
```



```{r}
#extract stops 
stops_starting <- stops_sf[stops_sf$intersect == 1,]
stops_destination <- stops_sf[stops_sf$stop_id %in% unique(dist_sf$to_stop),]
rm(dist_sf0,stops_sf)

#clean up in the amount of information in the dataframes 
dist_sf <- dist_sf[1:7]

map <- mapview(list(EDU, dist_sf, stops_starting, stops_destination), 
        layer.name= c("Programmes", "Travel Paths", "Stops (start)", "Stops (dest)"), 
        burst = T, 
        zcol = list("STATUS", "limit", NULL, NULL))
##save map 
mapshot(map, url = "../out/PTARmap.html", selfcontained = F)
```

