---
title: "Preprocess Rejseplanen LABS data"
output: html_document
date: "2024-05-30"
---

Data aquisition and pre-processing is described in detail in the README file, and the main points are included in the exam write up. The GTFS data is used to estimate the difference in acces and reach of public transporting from different locatioons. 

# Set Up and Package Installation
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load("tidyverse", "sf")
```


# Load the sf object with the starting point locations and add Buffer
```{r}
EDU <- st_read("../in/preprocessed/education/EDU.shp") #load in EDU data
EDU_BUF <- EDU #duplicate 
EDU_BUF["geometry"] <- st_buffer(EDU[c("geometry")], 500) #add 500 meter buffer to it
```

# Extract and Pre-Process the raw GTFS data
Before creating variables representing distances and time etc, the raw data is pre-processed. The data set is very large so the aim is to extract all relevant information and remove the rest, to not over-load the system. 

## routes.txt
```{r}
routes <- read.csv("../in/GTFS/routes.txt")#load in route data; info on all routes
routes <- routes %>% 
  filter(!agency_id == 401) %>% #remove the routes with agency Skånetrafikken
  filter(!route_type %in% c(715,4)) %>%  #remove flextrafik and ferries
  dplyr::select("route_id", "agency_id", "route_type")
```

## trips.txt
```{r}
trips <- read.csv("../in/GTFS/trips.txt") #load in trip data; info on each unique trip
trips <- trips %>% 
  dplyr::select("route_id", "service_id", "trip_id", "direction_id", "shape_id") %>% 
  filter(route_id %in% routes$route_id) #extract trips belonging to routes from routes
rm(routes) #routes is no longer needed so it is removed 
```

## stop_times.txt
```{r}
stop_times <- read_csv("../in/GTFS/stop_times.txt", #load stop times with the columns: 
                       col_select = c("trip_id", #id for unique trip 
                                      "departure_time", #time of departure(arrival removed)
                                      "stop_id", #id for the given stop
                                      "stop_sequence"), #the number stop in the trip
                       col_types = cols("n", "c", "t","n", "i","n", "n", "c"))

stop_times <- stop_times %>% filter(trip_id %in% trips$trip_id) #only trips from trips are kept

## departure_times after midnight in a trip that starts before midnight have values above 24; 25:01 is 01:01. Most datetime classes cannot handle this, so these are changed to a readable format (nothing above 24:00).
for (i in 1:nrow(stop_times)){
  timestamp <- as.numeric(strsplit(as.character(stop_times$departure_time[i]), ":")[[1]])
  if (timestamp[1]>23){
    timestamp[1] <- timestamp[1]-24
    timestamp <- str_flatten(as.character(timestamp), collapse = ":")
    timestamp <- hms::as_hms(timestamp) 
    stop_times$departure_time[i] <- timestamp
  }
}

stop_times <- stop_times %>% 
  mutate(time = departure_time-lag(departure_time)) #the time col represents the seconds it took from the previous stop to the current stop
stop_times$time[stop_times$stop_sequence == 0] <- 0 #to seperate different trips every seq 0 is assigned 0 seconds. 

stop_times$time[stop_times$time<0] <- stop_times$time[stop_times$time<0] +86400 #mesuring time across midnight is tricky; to retireve the actual number of seconds, the number of seconds in 24 hours are added times below 0. 

stop_times <- merge(stop_times, trips) #merge with trips 
stops_routes <- unique(stop_times[c(3:9)]) #unique rows with stop_id, stop_sequence, route_id, time, service_id, direction_id, shape_id
write_csv(stops_routes, "../in/preprocessed/GTFS_prep/stops_routes.csv")
write_csv(stop_times, "../in/preprocessed/GTFS_prep/stop_times.csv")
rm(trips, i, timestamp)#trips is no longer relevant 
```


## stops.txt
```{r}
stops_sf <- read.csv("../in/GTFS/stops.txt") #load in stops.txt 
stops_sf <- stops_sf %>% 
  st_as_sf(coords = c("stop_lon", "stop_lat"), crs = 4326) %>% #convert to sf with ESPG 4326
  st_transform(crs = 25832) %>%  #transform to planar coordinates ESPG: 25832
  select("stop_id", "stop_name") #keep stop_id, stop_name and geometry column 
#to be able distinct between stops within the 500 meter radius from the programmes and intersect column is added to stops_sf. 0 for stops not intersecting, 1 for stops that do. 
stops_sf$intersect <- 0
stops_sf$intersect[c(unique(unlist(st_intersects(EDU_BUF, stops_sf))))] <- 1 
st_write(stops_sf, "../in/preprocessed/GTFS_prep/stops.shp", append = FALSE)
```

## shapes.txt
```{r}
shapes <- read.csv("../in/GTFS/shapes.txt", fileEncoding = "latin1") #load in shapes.txt
shapes_sf <- shapes %>% 
  st_as_sf(coords = c("shape_pt_lon", "shape_pt_lat"), crs = 4326) %>% #convert to sf 
  st_transform(crs = 25832) #transform to planar with ESPG: 25832
##the shapes are in POINT format, these are casted to LINESTRING, to create one row per shape with a geometry column and a shape_id
shapes_sf <- shapes_sf %>% 
  group_by(shape_id) %>%
  dplyr::summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")
st_write(shapes_sf, "../in/preprocessed/GTFS_prep/shapes.shp", append = F)
rm(shapes)
```

## calendar.txt
```{r}
calendar <- read_csv("../in/GTFS/calendar.txt", #load in calendar.txt
                     col_types = cols("i", "i", "i", "i", "i", "i", "i", "i", "d", "d"))
calendar <- calendar %>% 
  filter(service_id %in% stops_routes$service_id) #keep only service_id from the stop_routes
```

## calendar_dates.txt
```{r}
calendar_dates <- read_csv("../in/GTFS/calendar_dates.txt", #load in calendar_dates
                           col_types = cols("n", "d", "n"))
calendar_dates <- calendar_dates %>% 
  filter(service_id %in% stops_routes$service_id) %>% #keep only service ids in stop_routes
  mutate(date = as.Date(as.character(date), "%Y%m%d")) %>% #define the date column as a Date object
  mutate(service_id = as.character(service_id))#convert service id to character
```


# Reformat the GTFS Data

## Distance 


```{r}
stops_routes <- read_csv("../in/preprocessed/GTFS_prep/stops_routes.csv")
stops_sf <- st_read("../in/preprocessed/GTFS_prep/stops.shp")
shapes_sf <- st_read("../in/preprocessed/GTFS_prep/shapes.shp")
EDU <- st_read("../in/preprocessed/education/EDU.shp") #load in EDU data
EDU_BUF <- EDU #duplicate 
EDU_BUF["geometry"] <- st_buffer(EDU[c("geometry")], 500) #add 500 meter buffer to it
```


```{r}
columns <- c("EDU_i", "limit", "stop_id", "to_stop", "time", 
             "route_id", "serivce_id", "direction_id", "shape_id")
dist_df = data.frame(matrix(nrow = 0, ncol = length(columns))) 
colnames(dist_df) <- columns

timelimits <- c((5*60), (20*60), (45*60))

for (i in 1:nrow(EDU_BUF)){
  stops_intersect <- stops_sf[unlist(st_intersects(EDU_BUF[i,], stops_sf)),]$stop_id
  for (limit in timelimits){
    for (stop in stops_intersect){
      for (j in 1:nrow(stops_routes)){
        if (stops_routes$stop_id[j] == stop){
          dist_df[nrow(dist_df)+1,] <- c(i, 
                                            sprintf("%s minute", (limit/60)),
                                            stop, 
                                            NA, 
                                            0.0,
                                            (unname(as.vector(stops_routes[j, 4:7]))))
          for (k in ((j+1):nrow(stops_routes))){
            if (stops_routes$stop_sequence[k]>0){
              dist_df[nrow(dist_df), "to_stop"] <- stops_routes$stop_id[k]
              dist_df[nrow(dist_df), "time"] <- sum(dist_df[nrow(dist_df), "time"],
                                                    stops_routes$time[k])}
            else if (stops_routes$stop_sequence[k]==0){
              break }
            if (dist_df[nrow(dist_df),"time"]>= limit){
              break }
          }
        }
      }
    }
  }
}

dist_df <- dist_df %>% 
  filter(limit == "5 minute" & time <= 300 |
         limit == "20 minute" & time <= 1200 |
         limit == "45 minute" & time <= 2700)  ##filter out duplicates
## for all rows not able to reach another stop within the alloted time limits, the to_stop is set to stop_id (so when measuring distance = 0) 
dist_df$to_stop[is.na(dist_df$to_stop)] <- dist_df$stop_id[is.na(dist_df$to_stop)]

write_csv(dist_df, "../in/preprocessed/GTFS_prep/distancesALL.csv")
```




## distances continued
```{r}
dist_df$distance <- 0.0
shapes_edu <- shapes_sf[shapes_sf$shape_id %in% dist_df$shape_id ,]
dist_sf <- merge(shapes_edu, dist_df, all.y = T)

dist_function <- function(from_stopG, to_stopG, shapeG){
  shape_points <- st_cast(shapeG, "POINT")
  
  shape_sect <- shape_points[st_nearest_feature(to_stopG, 
                                                shape_points):st_nearest_feature(from_stopG, 
                                                                                 shape_points), ]
  sectionLINE <- shape_sect %>% 
    group_by(shape_id) %>%
    dplyr::summarise(do_union = FALSE) %>%
    st_cast("LINESTRING")
  return(sectionLINE$geometry)
}


for (i in 1:nrow(dist_sf)){
  from_stopG <- stops_sf$geometry[stops_sf$stop_id == dist_sf$stop_id[i],]
  to_stopG <- stops_sf$geometry[stops_sf$stop_id == dist_sf$to_stop[i],]
  shapeG <- dist_sf[i,c("shape_id", "geometry")]
  
  dist_sf$geometry[i] <- dist_function(from_stopG, to_stopG, shapeG)
  dist_sf$distance[i] <- st_length(dist_sf$geometry[i])
}
st_write(dist_sf, "../in/preprocessed/GTFS_prep/distancesALL.shp", append = F)
```







