---
title: "Preprocess Rejseplanen LABS data"
output: html_document
date: "2024-05-30"
---

Data aquisition and pre-processing is described in detail in the README file, and the main points are included in the exam write up. The GTFS data is used to estimate the difference in acces and reach of public transporting from different locatioons. 

# Set Up and Package Installation
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load("tidyverse", "sf")
```


# Load the sf object with the starting point locations and add Buffer
```{r}
EDU <- st_read("../in/preprocessed/education/EDU.shp") #load in EDU data
EDU_BUF <- EDU #duplicate 
EDU_BUF["geometry"] <- st_buffer(EDU[c("geometry")], 500) #add 500 meter buffer to it
```

# Extract and Pre-Process the raw GTFS data
Before creating variables representing distances and time etc, the raw data is pre-processed. The data set is very large so the aim is to extract all relevant information and remove the rest, to not over-load the system. 

## routes.txt
```{r}
routes <- read.csv("../in/GTFS/routes.txt")#load in route data; info on all routes
routes <- routes %>% 
  filter(!agency_id == 401) %>% #remove the routes with agency Skånetrafikken
  filter(!route_type %in% c(715,4)) %>%  #remove flextrafik and ferries
  dplyr::select("route_id", "agency_id", "route_type")
```

## trips.txt
```{r}
trips <- read.csv("../in/GTFS/trips.txt") #load in trip data; info on each unique trip
trips <- trips %>% 
  dplyr::select("route_id", "service_id", "trip_id", "direction_id", "shape_id") %>% 
  filter(route_id %in% routes$route_id) #extract trips belonging to routes from routes
rm(routes) #routes is no longer needed so it is removed 
```

## stop_times.txt
```{r}
stop_times <- read_csv("../in/GTFS/stop_times.txt", #load stop times with the columns: 
                       col_select = c("trip_id", #id for unique trip 
                                      "departure_time", #time of departure(arrival removed)
                                      "stop_id", #id for the given stop
                                      "stop_sequence"), #the number stop in the trip
                       col_types = cols("n", "c", "t","n", "i","n", "n", "c"))

stop_times <- stop_times %>% filter(trip_id %in% trips$trip_id) #only trips from trips are kept

## departure_times after midnight in a trip that starts before midnight have values above 24; 25:01 is 01:01. Most datetime classes cannot handle this, so these are changed to a readable format (nothing above 24:00).
for (i in 1:nrow(stop_times)){
  timestamp <- as.numeric(strsplit(as.character(stop_times$departure_time[i]), ":")[[1]])
  if (timestamp[1]>23){
    timestamp[1] <- timestamp[1]-24
    timestamp <- str_flatten(as.character(timestamp), collapse = ":")
    timestamp <- hms::as_hms(timestamp) 
    stop_times$departure_time[i] <- timestamp
  }
}

stop_times <- stop_times %>% 
  mutate(time = departure_time-lag(departure_time)) #the time col represents the seconds it took from the previous stop to the current stop
stop_times$time[stop_times$stop_sequence == 0] <- 0 #to seperate different trips every seq 0 is assigned 0 seconds. 

stop_times$time[stop_times$time<0] <- stop_times$time[stop_times$time<0] +86400 #mesuring time across midnight is tricky; to retireve the actual number of seconds, the number of seconds in 24 hours are added times below 0. 

stop_times <- merge(stop_times, trips) #merge with trips 
stops_routes <- unique(stop_times[c(3:9)]) #unique rows with stop_id, stop_sequence, route_id, time, service_id, direction_id, shape_id
write_csv(stops_routes, "../in/preprocessed/GTFS_prep/stops_routes.csv")
write_csv(stop_times, "../in/preprocessed/GTFS_prep/stop_times.csv")
rm(trips)#trips is no longer relevant 
```


## stops.txt
```{r}
stops_sf <- read.csv("../in/GTFS/stops.txt") #load in stops.txt 
stops_sf <- stops_sf %>% 
  st_as_sf(coords = c("stop_lon", "stop_lat"), crs = 4326) %>% #convert to sf with ESPG 4326
  st_transform(crs = 25832) %>%  #transform to planar coordinates ESPG: 25832
  select("stop_id", "stop_name") #keep stop_id, stop_name and geometry column 
#to be able distinct between stops within the 500 meter radius from the programmes and intersect column is added to stops_sf. 0 for stops not intersecting, 1 for stops that do. 
stops_sf$intersect <- 0
stops_sf$intersect[c(unique(unlist(st_intersects(EDU_BUF, stops_sf))))] <- 1 
st_write(stops_sf, "../in/preprocessed/GTFS_prep/stops.shp", append = FALSE)
```

## shapes.txt
```{r}
shapes <- read.csv("../in/GTFS/shapes.txt", fileEncoding = "latin1") #load in shapes.txt
shapes_sf <- shapes %>% 
  st_as_sf(coords = c("shape_pt_lon", "shape_pt_lat"), crs = 4326) %>% #convert to sf 
  st_transform(crs = 25832) #transform to planar with ESPG: 25832
##the shapes are in POINT format, these are casted to LINESTRING, to create one row per shape with a geometry column and a shape_id
shapes_sf <- shapes_sf %>% 
  group_by(shape_id) %>%
  dplyr::summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")
st_write(shapes_sf, "../in/preprocessed/GTFS_prep/shapes.shp", append = F)
rm(shapes)
```

## calendar.txt
```{r}
calendar <- read_csv("../in/GTFS/calendar.txt", #load in calendar.txt
                     col_types = cols("i", "i", "i", "i", "i", "i", "i", "i", "d", "d"))
calendar <- calendar %>% 
  filter(service_id %in% stops_routes$service_id) #keep only service_id from the stop_routes
```

## calendar_dates.txt
```{r}
calendar_dates <- read_csv("../in/GTFS/calendar_dates.txt", #load in calendar_dates
                           col_types = cols("n", "d", "n"))
calendar_dates <- calendar_dates %>% 
  filter(service_id %in% stops_routes$service_id) %>% #keep only service ids in stop_routes
  mutate(date = as.Date(as.character(date), "%Y%m%d")) %>% #define the date column as a Date object
  mutate(service_id = as.character(service_id))#convert service id to character
```


# Reformat the GTFS Data
Using the pre-processed GTFS data, information on service dates, waiting time and distances are retireved. 
## calendar_matrix
```{r}
#define the column names, i.e. the date range based on calendar start and end date. 
date_span <- seq(as.Date(as.character(unique(calendar$start_date)), "%Y%m%d"), 
                 as.Date(as.character(unique(calendar$end_date)), "%Y%m%d"), by = "day")
#create empty matrix service_id as rownames and dates as columnames
calendar_matrix <- matrix(nrow = length(unique(calendar$service_id)), 
                     ncol = length(date_span), 
                     dimnames = list(calendar$service_id, as.character(date_span)))
#fill the matrix based on calendar
for (i in 1:nrow(calendar)){
  for (j in seq(1:7)){
    seq <- c(seq(from = j, to = ncol(calendar_matrix),  by = 7))
                 #,length.out = ncol(calendar_matrix)%/%7))
    if (calendar[i, j+1] == 1){
      calendar_matrix[i, seq] <- 1
    }
    else if (calendar[i, j+1] == 0){
      calendar_matrix[i, seq] <- 0
    }
  }
}
##fill the matrix based on calendar_date
for (i in 1:nrow(calendar_dates)){
  if (calendar_dates$exception_type[i] == 1){
    calendar_matrix[(calendar_dates$service_id[i]), 
                    as.character(calendar_dates$date[i])] <- 1
  }
  else if (calendar_dates$exception_type[i] == 2){
    calendar_matrix[(calendar_dates$service_id[i]), 
                    as.character(calendar_dates$date[i])] <- 0
  }
}
save(calendar_matrix, file = "../in/preprocessed/GTFS_prep/calendar_matrix.RData")
rm(calendar, calendar_dates)
```


## wait_time
The waiting times for the stops within the buffers are extracted, to add an average to the time limits when measuring distance. 
```{r}
stop_waits <- stop_times %>% 
  filter(stop_id %in%c(stops_sf$stop_id[stops_sf$intersect == 1])) %>% #keep only stops that intersect with the buffers
  select("trip_id", "departure_time", "stop_id", "route_id", "service_id", "shape_id") %>% 
  arrange(stop_id, service_id, departure_time) %>% #arrange so departure times are collected by stops and service_id
  group_by(stop_id, service_id) %>% #group by stop and service id 
  mutate(wait = departure_time-lag(departure_time))#create wait column with wait time
#to be able to handle each 24 hours as a 'loop' the last stop of the day has to be related back to the first stop of the 'next' day (meaning the same day). 
midnight <- stop_waits %>% 
  slice(c(1,n())) %>% #extract first and last stop times
  arrange(stop_id, service_id, desc(departure_time)) %>%  #arrange departure in decending order, as the wait is calculated by subtracting the previous time. 
  mutate(wait = (departure_time-lag(departure_time)+86400)) %>% #calculate wait, and add 86400 (n seconds in 24 hours) to retrieve the actual wait in seconds
  filter(row_number() %% 2 == 0) %>% #extract only every other row, the first stop of each stopxservice combination. 
  ungroup()
stop_waits <- stop_waits %>% 
  slice(-c(1)) %>% #remove the first row of every group
  ungroup()
stop_waits <- rbind(stop_waits, midnight) #combine the first rows back to the df 
write_csv(stop_waits, "../in/preprocessed/GTFS_prep/stop_waits.csv")
rm(midnight)
```




### Aggregate stop_waits
```{r}
stops_sf <- st_read("../in/preprocessed/GTFS_prep/stops.shp")
stop_times <- read_csv("../in/preprocessed/GTFS_prep/stop_times.csv")
stop_waits <- read_csv("../in/preprocessed/GTFS_prep/stop_waits.csv")
shapes_sf <- st_read("../in/preprocessed/GTFS_prep/shapes.shp")
stops_routes <- read_csv("../in/preprocessed/GTFS_prep/stops_routes.csv")

load("../in/preprocessed/GTFS_prep/calendar_matrix.RData")
```

##collapsing service ids 
```{r}
agg_waits <- stop_waits %>% 
  dplyr::summarise(mean_wait = mean(wait), 
                   sd_wait = sd(wait), 
                   n_group = n(),
                   .by = c(stop_id, service_id))

agg_waits$weekday <- 0
agg_waits$weekend <- 0


for ()


table((is.na(calendar_matrix)))


rowSums(calendar_matrix)

weekdays(as.POSIXlt(colnames(calendar_matrix)))


```




```{r}


stop_waits[stop_waits$wait ==86400,]
stop_times[stop_times$trip_id == 127803281,]



stop_waits[stop_waits$trip_id == 127803281,]



mapview(shapes_sf[shapes_sf$shape_id == 5484,]) + mapview(stops_sf[stops_sf$stop_id %in%stop_waits$stop_id[stop_waits$trip_id == 127803281],])
```

```{r}
library(mapview)


mapview(stops_sf[stops_sf$intersect == 1,])+mapview()

```





## time_limit
```{r}

```



# Functions 
## time_function(): 

```{r}
time_function <- function(edu_indx, stop_idx, stops_routes, time_df, thresh){
  
  for (i in (1:nrow(stops_routes))){
    if (stops_routes$stop_id[i] == stop_idx){
      time_df[nrow(time_df)+1,] <- list(edu_indx, 
                                        stop_idx, 
                                        NA, 
                                        0.0,
                                        sprintf("%s minute", (thresh/60)),
                                        0.0, 
                                        stops_routes[i,"route_id"], 
                                        stops_routes[i,"service_id"],
                                        stops_routes[i,"direction_id"],
                                        stops_routes[i,"shape_id"])
      for (j in ((i+1):nrow(stops_routes))){
        if (stops_routes$stop_sequence[j]>0){
          time_df[nrow(time_df),"to_stop"] <- stops_routes$stop_id[j]
          time_df[nrow(time_df),"time"] <- sum(time_df[nrow(time_df),"time"], stops_routes$time[j])
        }
        
        else if (stops_routes$stop_sequence[j]==0){
          break
        }
        
        if (time_df[nrow(time_df),"time"]>=min){
          break
        }
      }
    }
  }
  return(time_df)
}
```


#### time_all_fun():
- arguments: 
  - dataframe with one row per location (polygon) 
  - stop sf
  - list of time threholds in seconds
  - stop secs 
- creates the time_df dataframe
- extracts the intersection-stops (locations within polygons from stop_sf)

```{r}
time_all_fun <- function(EDU_BUF, stops_sf, thresholds, stop_secs){
  
  columns <- c("edu_indx", "from_stop", "to_stop", "time","min", "distance", 
               "route_id", "serivce_id", "direction_id", "shape_id")
  time_df = data.frame(matrix(nrow = 0, ncol = length(columns))) 
  colnames(time_df) <- columns
  stops_inter <- st_intersects(EDU_BUF, stops_sf)
  
  for (edu_indx in 1:nrow(EDU_BUF[1,])){
    stops = stops_sf[stops_inter[[edu_indx]],]$stop_id
    for (thresh in thresholds){
      for (i in 1:length(stops)){
        time_df = time_function(edu_indx, stops[i], stop_secs, time_df, thresh)
      }
    }
  }
  time_df <- time_df %>% 
  filter(
    min == "5 minute" & time <= 300 |
    min == "20 minute" & time <= 1200 |
    min == "45 minute" & time <= 2700) %>% 
  filter(!is.na(to_stop))
  return(time_df)
}
```


### Test functions 
```{r}
thresholds <- c((5*60), (20*60), (45*60))
time_df <- time_all_fun(EDU_BUF, stops_sf, thresholds, stop_secs)
```


```{r}
st_write(time_df, "../data/data_post/time_df.shp", append = F)
```



### Extracting Travel Time 
The possible travel times within the three thresholds; 5, 20 and 45 minutes are extracted for all stops within the 500 meter buffer of each institution. 
- First all stops within the 500 meter buffer are extracted 
- The the time_df edited for every instituion and stop. 
- The time_df is megeed with shapes_edu (all shapes used in the inst data)



#### Cut shapes to fit between designated points 
```{r}
shapes_edu <- shapes_sf[shapes_sf$shape_id %in% time_df$shape_id ,]
```

#### Clean up
- When checking whether the summed time is over or equal to the threshold some inaccurate data is being passed on, this is cleaned up; making sure that nothing is above the three threholds
- Additionally, some stops within the 500 m buffer is the end-station, they will have NA in their 'to_stop' column, and are deleted. 
- lastly the relevant time geometries are included
```{r}
time_df <- time_df %>% 
  filter(
    min == "5 minute" & time <= 300 |
    min == "20 minute" & time <= 1200 |
    min == "45 minute" & time <= 2700) %>% 
  filter(!is.na(to_stop))
time_df <- merge(shapes_edu, time_df, all.y = T)
```




### Extracting Distance between all from_stops and to_stops

consider handeling a shape at a time instead of a row; to limit overhead
```{r}
##overall function extracts stop and shape geometries passed directly? 

gen_dist_fun <- function(shapes_edu, stops_sf, time_df){
  
  #for row in time df here 
  for (i in 1:nrow(time_df)){
    from_stopG <- stops_sf$geometry[stops_sf$stop_id == time_df$from_stop[i],]
    to_stopG <- stops_sf$geometry[stops_sf$stop_id == time_df$to_stop[i],]
    shapeG <- shapes_edu$geometry[shapes_edu$shape_id == time_df$shape_id[i],]
    time_df$distance[i] <- dist_function(from_stopG, to_stopG, shapeG)
  }
}
```




## EDU 
The EDU data frame has been preprocessed in preprocess_inst_edu.rmd and contains all new educations and their old counterparts as well as their locations in POINT geometry. A new sf object is made, with polygons instead, representing a 500 meter buffer around each location.
```{r}
EDU <- st_read("../in/preprocessed/EDU.shp")
EDU_BUF <- EDU
EDU_BUF["geometry"] <- st_buffer(EDU[c("geometry")], 500)
```



```{r}
dist_function <- function(from_stopG, to_stopG, shapeG){
  shape_points <- st_cast(shapeG, "POINT")
  shape_sect <- shape_points[st_nearest_feature(to_stopG, 
                                                shape_points):st_nearest_feature(from_stopG, 
                                                                                 shape_points), ]
  
  sectionLINE <- shape_sect %>% 
    group_by(shape_id) %>%
    dplyr::summarise(do_union = FALSE) %>%
    st_cast("LINESTRING")
  
  return(sectionLINE$geometry)
}
```

```{r}
##test dist fun 
for (i in 1:nrow(time_df)){
  from_stopG <- stops_sf$geometry[stops_sf$stop_id == time_df$from_stop[i],]
  to_stopG <- stops_sf$geometry[stops_sf$stop_id == time_df$to_stop[i],]
  shapeG <- time_df[i,c("shape_id", "geometry")]

  time_df$geometry[i] <- dist_function(from_stopG, to_stopG, shapeG)
  time_df$distance[i] <- st_length(time_df$geometry[i])
}
```


```{r}
mapview(time_df, zcol = "min", burst = T)
```


